Subset coercions in \Coq

### 2006-05-01 papers/article/main.tex
Typesetting fixes

\usepackage{bussproofs}
% Optional to turn on the short abbreviations
\EnableBpAbbreviations

@@

\input{typing-algo}

@@

\subsubsection{The typing judgment}
\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the \CCfull \cite{Barras99,Luo90}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is convertible to $U$ in environment $`G$.
Following the presentation as a pure type system \cite{PTS}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$, the
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
rules $\PTSrules$ are defined by the following (functional) relation:
\axiomsd

@@

Some basic properties of this system are well-formedness of
environments: if $`G \typed t : T$ then $\typedwf `G$ and
well-sortedness: if $`G \typed t : T$ then $T = \Type$ or $`G \typed T :
s$ for $s `: \PTSsorts$.

@@

\begin{figure*}[h]
  \typedRules
  \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
\end{figure*}

@@

The really interesting rules are \irule{SubSub} and
\irule{SubProof}. The first one allows to use an object of a subset type
as an object of its support type. The latter allows (maybe abusively) to
consider an object of any type as an object of any subset based on
it. By the assumption on the input types of the judgment, we know that $P$
is well-typed, but we do not care about the proof.

@@

We write $`G \typed T, U : s$ for $`G \typed T : s$ and $`G \typed U :
s$.

@@

  By induction on the subtyping derivation:
  \case{SubConv} We have $T \eqbp U$, we apply unicity of sorting.

  \case{SubTrans} We have by induction $`G \typed T \sub S : s$ and 
  $`G \typed S \sub U : s$. Hence we have $`G \typed T, U : s$.

  \caseother{} Other cases are straightforward.
\end{proof}

We will omit $`G$ where unnecessary.
The essential properties of the $\_ \typed \_ \subd \_$ judgment are 
its reflexivity ($\forall A, A \subd A$), symmetry ($\forall A B, A \subd
B "<=>" B \subd A$) and transitivity by \irule{SubTrans}.

@@

\subsubsection{Algorithmic typing}
We use a standard transformation to get an algorithm from our typing
rules, basically restricting the conversion (or subsumption) rule
(\irule{Coerce}) to the premises of introduction and elimination rules.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.

>>>>

\usepackage{bussproofs}
% Optional to turn on the short abbreviations
\EnableBpAbbreviations
\def\ScoreOverhang{3pt}			% How much underlines extend out
\def\defaultHypSeparation{\hskip.05in}   % Used if \insertBetweenHyps isn't given

@@

\input{typing-algo}
\renewcommand{\AppAFull}[9][\Gamma]{
\TAX{App}
{$\tgen{#1}{#2}{#3} \suba \Pi #4 : #5. #6$}
{$\tgen{#1}{#8}{#9}$} % \quad \tgen{#1}{#9, #5}{#7'}$}
{$\subalgo{#1}{#9}{#5}$}
{$\tgen{#1}{(#2 #8)}{#6 [ #8 / #4 ]}$}
{}
}
\input{subtyping-algo}
\renewcommand{\SubSubAs}[1][\Gamma]{%
\UAX{SubSub}
{$\subalgoimpl{#1}{c}{U}{T}$}
{$\subalgoimpl{#1}{c[\eltpit~\ctxdot]}{\mysubset{x}{U}{P}}{T}$}
{}
}

@@

\subsubsection{The typing judgment}

\begin{figure*}[t]
    \def\seq{\typed}
    \def\fCenter{\wf}
    \begin{center}
      \WfEmpty\DP\WfVar\DP
      \Var\DP
    \end{center}

    \def\fCenter{\typed}
    \begin{center}
      \PropSet\DP
      \Subsum\DP
    \end{center}

    \Prod\DP
    \SigmaR\DP

    \vspace{\infvspace}
    \Abs\DP
    \SumDep\DP

    \vspace{\infvspace}
    \quad\App\DP
    \quad\quad
    \PiLeft\DP
    \quad
    \PiRight\DP

    \begin{center}\SubsetR\DP\end{center}
    \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
\end{figure*}

\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the Calculus of Constructions \cite{CoquandHuet,Barras99}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is convertible to $U$ in environment $`G$. 
Following the presentation as a pure type system \cite{PTS}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$. As usual, we
let $s, s_i$ for $i `: `N$ range over sorts. The
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
rules $\PTSrules$ are defined by the following (functional) relation:
\begin{table}[h]
  \vspace{-1.5em}
  \axiomsd
  \vspace{-2em}
  \caption{Definition of $\PTSrules$}
  \label{fig:PTSrules}
\end{table}

@@

Some basic properties of this system are well-formedness of
environments: if $`G \typed t : T$ then $\typedwf `G$ and
well-sortedness: if $`G \typed t : T$ then $T = \Type$ or $`G \typed T :
s$.

@@

@@

The really interesting rules are \irule{SubSub} and
\irule{SubProof}. The first one allows to use an object of a subset type
as an object of its support type. The latter allows (maybe abusively) to
consider an object of any type as an object of any subset based on
this type. By the assumption on the input types of the judgment, we know
that the property $P$ is well-formed, but we do not care about the proof.

@@

\begin{paragraph}{Notations}
  We write $`G \typed T, U : s$ for $`G \typed T : s$ and $`G \typed U :
  s$.
\end{paragraph}

@@

  By induction on the subtyping derivation:
  \begin{list}{}
  \item[] \rulename{SubConv}: We have $T \eqbp U$, we apply unicity of sorting.

  \item[] \rulename{SubTrans}: We have by induction $`G \typed T \sub S : s$ and 
    $`G \typed S \sub U : s$. Hence we have $`G \typed T, U : s$.

  \item[] Other cases are straightforward.
  \end{list}
\end{proof}

The essential properties of the $\subd$ judgement are 
its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
B \subd A$ and transitivity by \irule{SubTrans}.

@@

\subsubsection{Algorithmic typing}
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}

    \begin{center}
    \WfEmpty\DP

    \vspace{\infvspace}
    \WfVar\DP

    \vspace{\infvspace}
    \Var\DP

    \vspace{\infvspace}
    \PropSet\DP

    \vspace{\infvspace}
    \Prod\DP

    \vspace{\infvspace}
    \Abs\DP

    \vspace{\infvspace}
    \AppA\DP

    \vspace{\infvspace}
    \SigmaR\DP

    \vspace{\infvspace}
    \SumDepA\DP

    \vspace{\infvspace}
    \PiLeftA\DP

    \vspace{\infvspace}
    \PiRightA\DP

    \SubsetR\DP
    \end{center}
  \caption{\Russell{} algorithmic typing and equivalence}
  \label{fig:typing-algo-rules}
\end{figure*}

\begin{figure*}
    \def\fCenter{\suba}
    \def\type{\typea}
    \def\sub{\suba}
    \begin{center}
    \SubConvA\DP

    \vspace{\infvspace}
    \SubHnfA\DP

    \vspace{\infvspace}
    \SubProdA\DP
    \quad
    \SubSigmaA\DP

    \vspace{\infvspace}
    \SubProofA\DP
    \quad
    \SubSubAs\DP
  \end{center}
  \caption{\Russell{} algorithmic typing and equivalence}
  \label{fig:subtyping-algo-rules}
\end{figure*}

We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
$T `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.
